---
title: "Forecasting"
author: "Maurício Collaça"
date: "15 de agosto de 2018"
output: 
  html_document: 
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
        collapsed: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment="")
```

## Forecasting

* What forecasting methods are available that take account of trend, seasonality and other features of the data?  
* How to measure the accuracy of your forecasts?  
* How to choose a good forecasting model?

## Creating time series objects in R

```{r}
library(readxl)
mydata <- read_excel("exercise1.xlsx")
```

```{r}
myts <- ts(mydata[c("Sales", "AdBudget", "GDP")], start = c(1981, 1), frequency = 4)
window(myts, end=c(1982,2))
```

# Time series Plots
```{r}
library(forecast)
autoplot(myts, facets=TRUE)
```
```{r}
autoplot(myts, facets=FALSE)
```

```{r}
autoplot(gold)
autoplot(woolyrnq)
autoplot(gas)
```

```{r}
frequency(gold)
frequency(woolyrnq)
frequency(gas)
```

# Seasonal plots

```{r}
library(fpp2)
frequency(a10)
autoplot(a10)
ggseasonplot(a10)
ggseasonplot(a10, polar = TRUE)
```

```{r}
beer <- window(ausbeer, start=c(1992,1))
frequency(beer)
autoplot(beer)
ggseasonplot(beer)
ggseasonplot(beer, polar=TRUE)
ggsubseriesplot(beer)
```

# Time series patterns

Pattern  | Description
---------|-----------------------------------------------------------------------
Trend    | A pattern exists involving a long-term increase OR decrease in the data (this is deliberaly a little vague because a trend is not a well-defined mathematical term)
Seasonal | A periodic pa!ern exists due to the calendar (e.g. the quarter, month, or day of the week)
Cyclic   | A pattern exists where the data exhibits rises and falls that are not of fixed period (duration usually of at least 2 years)

```{r}

```

## Examples of time series patterns

Trended and seasonal
```{r}
autoplot(window(qauselec, start=1980, end=1995))
```

Trended, Seasonal and Cyclic

```{r}
autoplot(bricksq)
```

No seasonality, looks like a downward trend but it's actually a part of a much longer cycle ( part of a cycle can look like a trend)

```{r}
autoplot(ustreas)
```

Because it's annual, it can be seasonal.  The lynx population rises when there's plenty of food and stoop breeding when lacking food.  The lenght of this cycle is between 8 to eleven years
```{r}
autoplot(lynx)
```

# Seasonal or cyclic?

Differences between seasonal and cyclic patterns:

* Seasonal pattern constant length vs. cyclic pattern variable length  
* If both exists together, the average length of cycle pattern is longer than length of seasonal pa!ern  
* The Magnitude (size) of the cycles tend to be more variable than the magnitude (size) of seasonal pattern

The timing of peaks and troughs is predictable with seasonal data, but unpredictable in the long term with cyclic data.

# Autocorrelation of non-seasonal time series

autoplot of the oil data
```{r}
str(oil)
```

```{r}
autoplot(oil) + geom_point()
```

Create a lag plot of the oil data
```{r}
gglagplot(oil, lags=12, do.lines = F) + geom_smooth(method=lm, color="red")
```

Create an ACF plot of the oil data
```{r}
ggAcf(oil, lag.max = Inf)
```

# Autocorrelation of seasonal and cyclic time series

autoplot the annual sunspot numbers
```{r}
str(sunspot.year)
```
```{r}
autoplot(sunspot.year)# + geom_point()
```

Create a lag plot of the annual sunspot numbers
```{r}
gglagplot(sunspot.year, lags=12, do.lines = F) + geom_smooth(method=lm, color="red")
```

```{r}
ggAcf(sunspot.year)
```

Save the lag corresponding to maximum autocorrelation
```{r}
frequency(sunspot.year)
(maxlag_sunspot <- 1)
```

autoplot the traffic on the Hyndsight blog
```{r}
str(hyndsight)
```
```{r}
autoplot(hyndsight)# + geom_point()
```

Create a lag plot of the traffic on the Hyndsight blog
```{r}
gglagplot(hyndsight, lags=12, do.lines = F) + geom_smooth(method=lm, color="red")
```

```{r}
ggAcf(hyndsight)
```

Save the lag corresponding to maximum autocorrelation
```{r}
frequency(hyndsight)
(maxlag_hyndsight <- 7)
```

# Hints

* Trends induce positive correlations in the early lags.  
* Seasonality will induce peaks at the seasonal lags.  
* Cyclicity induces peaks at the average cycle length.

# White Noise

"White noise" is just a time series of iid data which can't be used alone to build a forecasting model.
```{r}
wn <- ts(rnorm(36))
autoplot(wn)
```

* In WN it's expected each correlation is close to zero  
* 95% of all correlations for WN should lie between the blue lines, otherwise series is probably not WN.
```{r}
ggAcf(wn) + ggtitle("Sample ACF for white noise")
```

# Example: Pigs slaughtered

At first glance it looks relatively random or possibly a slightly upward trend, but it's hard to see the difference from WN.
```{r}
pigs <- window(pigs, start=1990)
autoplot(pigs/1000) +
    xlab("Year") +
    ylab("thousands") +
    ggtitle("Monthly number of pigs slaughtered in Victoria after 1990")
```

ACF show the first 3 spikes are significantly larger than zero confirming it's not WN.  There is some information in data that can be used in building a forecasting model.
```{r}
ggAcf(pigs) + ggtitle("ACF of monthly pigs slaughtered in Victoria")
```

# Ljung-Box test

Looking at the ACF is usefull but sometimes is easier to test all the autocorrelations together, rather then considering one separately.

The Ljung-Box test considers the first h autocorrelation values together.

A significant test (small p-value) indicates the data are probably not white noise.

Applying it to the first 24 autocorrelation in the pigs dataset.
```{r}
Box.test(pigs, lag = 24, fitdf = 0, type = "Ljung-Box")
```
The p-value is very small, thus rejecting the null hypothesis of an independent time series, than suggesting this is not WN.

# White noise summary

* White noise is a time series that is purely random  
* We can test for white noise by looking at an ACF plot or by doing a Ljung-Box test

# Stock prices and white noise

As you learned in the video, white noise is a term that describes purely random data. You can conduct a Ljung-Box test using the function below to confirm the randomness of a series; a p-value greater than 0.05 suggests that the data are not significantly different from white noise.

There is a well-known result in economics called the **"Efficient Market Hypothesis"** that states that asset prices reflect all available information. A consequence of this is that the **daily changes in stock prices should behave like white noise (ignoring dividends, interest rates and transaction costs)**. The consequence for forecasters is that **the best forecast of the future price is the current price**.

You can test this hypothesis by looking at the goog series.

## Quandl data

```{r}
library(Quandl)
Quandl.api_key(readLines("Quandl.api_key.txt",1))
goog <- Quandl("WIKI/GOOG", type="xts", start_date="1990-01-01")
str(goog)
goog <- goog["/2017-02-13"]
head(goog[, c("Close", "Adj. Close")])
tail(goog[, c("Close", "Adj. Close")])
```

Plot the original series
```{r}
autoplot(goog[, c("Close","Adj. Close")])
```

Plot the differenced series
```{r}
autoplot(diff(goog[, c("Close","Adj. Close")]))
```

ACF of the differenced series
```{r}
ggAcf(diff(goog[, "Close"]))
ggAcf(diff(goog[, "Adj. Close"]))
```

Ljung-Box test of the differenced series
```{r}
Box.test(diff(goog[, "Close"]), lag = 10, type = "Ljung")
Box.test(diff(goog[, "Adj. Close"]), lag = 10, type = "Ljung")
```

## quantmod data
```{r}
library(quantmod)
getSymbols("GOOG")
str(GOOG)
GOOG <- GOOG["/2017-02-13"]
head(GOOG[, c("GOOG.Close", "GOOG.Adjusted")])
tail(GOOG[, c("GOOG.Close", "GOOG.Adjusted")])
```

Plot the original series
```{r}
autoplot(GOOG[, c("GOOG.Close", "GOOG.Adjusted")])
```

Plot the differenced series
```{r}
autoplot(diff(GOOG[, c("GOOG.Close", "GOOG.Adjusted")]))
```

ACF of the differenced series
```{r}
ggAcf(diff(GOOG[, "GOOG.Close"]))
ggAcf(diff(GOOG[, "GOOG.Adjusted"]))
```

Ljung-Box test of the differenced series
```{r}
Box.test(diff(GOOG[, "GOOG.Close"]), lag = 10, type = "Ljung")
Box.test(diff(GOOG[, "GOOG.Adjusted"]), lag = 10, type = "Ljung")
```

